A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-19 tex/codatadp.tex
introduces \Var notation and made it consistent with \Val (although I'm open-minded here)

%\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\vec}[1]{\bar #1}
\newcommand{\Ec}{\mathsf{E}}
\newcommand{\Fc}{\mathsf{F}}
\newcommand{\Gc}{\mathsf{G}}
\newcommand{\Ac}{\mathsf{A}}
\newcommand{\Dc}{\mathsf{D}}
\newcommand{\Rc}{\mathsf{R}}
\newcommand{\vrange}{\mathsf{range}}
\newcommand{\vdom}{\mathsf{dom}}
\newcommand{\tEc}{\mathcal{T}(\Ec)}
\newcommand{\tcEc}{\mathcal{T}^\ast(\Ec)}
\newcommand{\rn}[1]{\textsf{\small #1}}
\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\tneq}{\not\teq}
\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
%\newcommand{\is}[1]{is\text{-} #1}
\newcommand{\is}[1]{\const{is#1}}
\newcommand{\ror}{\quad \parallel \quad}
\newcommand{\tpath}[2]{\mathcal{P}_{ #2 \rightarrow \_ }( #1 )}
\newcommand{\ttpath}[3]{\mathcal{P}_{ #2 \rightarrow #3 }( #1 )}
\newcommand{\interp}[1]{[\![ #1 ]\!]}
\newcommand{\dpath}[3]{\delta^{#1}_{#2}( #3 )}
\newcommand{\ec}[1]{ [ #1 ] }
\newcommand{\M}{\mathcal{M}}
\newcommand{\Val}{\mathcal{A}}
\newcommand{\nf}[1]{{#1}\!\!\downarrow}
\newcommand{\vsim}{\sim}
\newcommand{\aequiv}{\mathrel{=_\alpha}}
\newcommand{\vsimv}[1]{\vsim_{#1}}
\newcommand{\tpos}[2]{#1\!\mid_{#2}}
\newcommand{\muvar}{\mathcal{Var}}

@@

\paragraph{Phase 2: Applying Acyclicity and Bisimilarity.}
%For presentation of the rules in this phase, we rely on a representation of terms in the $\mu$-notation.
The rules of the second phase contain premises that involve a mapping $\Val$
that associates a $\mu$-term as representative with each equivalence class.
%Recall that $\mu$-bindings can be used for representing possibly cyclic terms and values.
%(Section~\ref{sec:introduction}).
% for example, $\mu x. \const{C}( \const{0}, x )$ represents the cyclic value $\const{C}( \const{0}, \const{C}( \const{0}, \ldots ))$.
Intuitively, a $\mu$-term $t$ of type $\tau$ describes a class of values of $\tau$,
%either in the case when $\tau$ is a codatatype or datatype type,
where a cyclic $t$ describes an infeasible class of values when
%$\tau \not\in \Codata$.
$\tau \in \Data$.
Formally, $\mu$-terms are defined recursively as being either a variable $x$
or a node of the form
$\mu x.\; \const{C}( \vec t\, )$ for some constructor $\const{C} \in \Ctr$ and
$\mu$-terms $\vec t$ with the expected types.
The variable $x$ need not occur under the binder's body, in which case the
binder can be omitted.

@@

At the beginning of this phase, 
we compute a mapping $\Val$ from equivalence classes to $\mu$-terms 
that describe the class of values that terms in that equivalence class can take in the models of $\Ec$. 
For each equivalence class $\ec{x}$, we associate a fresh variable $v_{\ec{x}}$ of the same type as $x$ not occurring in $\tEc$,
and initially say that $\Val$ contains $\ec{x} \mapsto v_{\ec{x}}$,
in other words, there are initially no constraints on the values for any equivalence class $\ec{x}$.
Then, $\Val$ is modified by applying the following update exhaustively:
\[
\inferrule{
  v_{\ec{x}} \in \FV( \Val )
  \quad
  \const{C}( t_1, \ldots, t_n ) \in \ec{x}
  \quad
  \const{C} \in \Ctr
}{
  \Val := \Val \{ v_{\ec{x}} \mapsto \mu v_{\ec{x}}. \const{C}( v_{\ec{t_1}}, \ldots, v_{\ec{t_n}} ) \}
}
\]
We write $\FV( \Val )$ to denote the \relax{free variables} occurring in $\Val$, which
exclude variables that occur beneath $\mu$-binders.
It is straightforward to see that the height of terms produced as a result of this update
is bounded by the number of equivalence classes of $\Ec$,
and thus our construction of $\Val$ is terminating.

\begin{example}
Say that $\Ec$ contains (distinct) equivalence classes $\ec{x}$, $\ec{y}$, $\ec{z}$ and $\ec{w}$ 
where $\const{C}(z,x) \in \ec{y}$ and $\const{C}(w,y) \in \ec{x}$ for some $\const{C} \in \Ctr$.
A possible sequence of updates applied to $\Val$ is shown below.
In the first column, we show the equivalence class that was considered,
and in the second column, we show the content of $\Val$.
\[\begin{array}{|l|@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l|}
\hline
\text{Eqc} & \multicolumn{3}{c}{\Val} & \ 
\\
\hline
& \{ & \ec{x} \mapsto v_{\ec{x}}, & \ec{z} \mapsto v_{\ec{z}}, & \\
&    & \ec{y} \mapsto v_{\ec{y}}, & \ec{w} \mapsto v_{\ec{w}} & \} \\
\ec{x}
& \{ & \ec{x} \mapsto \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, v_{\ec{y}} ), & \ec{z} \mapsto v_{\ec{z}}, & \\
&    & \ec{y} \mapsto v_{\ec{y}}, & \ec{w} \mapsto v_{\ec{w}} & \} \\
\ec{y}
& \{ & \ec{x} \mapsto \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, \mu v_{\ec{y}}. \const{C}( v_{\ec{z}}, v_{\ec{x}} ) ), & \ec{z} \mapsto v_{\ec{z}}, & \\
&    & \ec{y} \mapsto \mu v_{\ec{y}}. \const{C}( v_{\ec{z}}, v_{\ec{x}} ), & \ec{w} \mapsto v_{\ec{w}} & \} \\
\ec{x}
& \{ & \ec{x} \mapsto \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, \mu v_{\ec{y}}. \const{C}( v_{\ec{z}}, v_{\ec{x}} ) ), & \ec{z} \mapsto v_{\ec{z}}, & \\
&    & \ec{y} \mapsto \mu v_{\ec{y}}. \const{C}( v_{\ec{z}}, \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, v_{\ec{y}} ) ), & \ec{w} \mapsto v_{\ec{w}} & \} \\
\hline
\end{array}\]
This indicates that the values for $x$ (resp. $y$) in models of $\Ec$
are of the form $\const{C}( v_{\ec{w}}, \const{C}( v_{\ec{z}}, \ldots ))$ (resp. $\const{C}( v_{\ec{z}}, \const{C}( v_{\ec{w}}, \ldots ))$) 
for some value of $v_{\ec{w}}$ and $v_{\ec{z}}$.
\qed
\end{example}

@@

\begin{example}
Let $\const{C}$ be a constructor of type $\tau_1 \times \tau_2 \rightarrow \tau_2$,
and let $\Ec$ be the set $\{ x \teq \const{C}(y,\const{C}(y,x)) \}$.
After applying the rules in Figure~\ref{fig:cc-rules} to completion, the equivalence classes of $\Ec$ are
$\{ x, \const{C}(y,\const{C}(y,x)) \}$, $\{ \const{C}(y,x) \}$, and $\{ y \}$.
We then construct $\Val$ using the aforementioned phases, after which we obtain:
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}r@{\hspace{.5em}}l@{\hspace{.5em}}r@{\hspace{.5em}}l}
\Val \ec{x} & = & 
\mu v_{\ec{x}}. & \const{C}( v_{\ec{y}}, & \mu v_{\ec{\const{C}(y,x)}}. & \const{C}( v_{\ec{y}}, v_{\ec{x}} ) ) \\
\Val \ec{\const{C}(y,x)} & = & 
\mu v_{\ec{\const{C}(y,x)}}. & \const{C}( v_{\ec{y}}, & \mu v_{\ec{x}}. & \const{C}( v_{\ec{y}}, v_{\ec{\const{C}(y,x)}} ) )
\end{array}\]
Notice that $\Val \ec{x}$ and $\Val \ec{\const{C}(y,x)}$ are syntactically equivalent for the renaming
$\{ v_{\ec{x}} \mapsto v_{\ec{\const{C}(y,x)}}, v_{\ec{\const{C}(y,x)}} \mapsto v_{\ec{x}} \}$,
and thus our calculus may infer $x \teq \const{C}(y,x)$.
Intuitively, this equality holds, since the values of $x$ and $\const{C}( y, x )$ 
are equivalent for any value of $v_{\ec{y}}$.
\qed
\end{example}

@@

\begin{example}
Let $\tau$ be a type in $\Codata$ with three constructors $\const{A}$, $\const{B}$ and $\const{C}$ each having type $\tau \rightarrow \tau$.
Let $\Ec$ be the set consisting of equalities
$\{$
$y_1 \teq \const{A}( y_2 )$,
$y_2 \teq \const{B}( x )$,
$z_1 \teq \const{C}( z_2 )$,
$z_2 \teq \const{A}( z_1 )$,
$w_1 \teq \const{A}( x )$,
$w_2 \teq \const{C}( w_3 )$,
$x \not\teq y_2$,
$\}$.
After running the calculus to saturation on $\Ec$, the mapping $\Val$ contains:
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{2.0em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
\Val \ec{y_1} & = & \mu v_{\ec{y_1}}. & \const{A}( & \mu v_{\ec{y_2}}. & \const{B}( v_{\ec{x}} ) ) &
\Val \ec{w_1} & = & \mu v_{\ec{w_1}}. & \const{A}( v_{\ec{x}} )      & \\ 
\Val \ec{y_2} & = & \mu v_{\ec{y_2}}. & \multicolumn{2}{l}{\const{B}( v_{\ec{x}} )}      & &
\Val \ec{w_2} & = & \mu v_{\ec{w_2}}. & \const{C}( v_{\ec{w_3}} ) \\
\Val \ec{z_1} & = & \mu v_{\ec{z_1}}. & \const{C}( & \mu v_{\ec{z_2}}. & \const{A}( v_{\ec{z_1}} ) ) &
\Val \ec{w_3} & = & v_{\ec{w_3}} \\
\Val \ec{z_2} & = & \mu v_{\ec{z_2}}. & \const{A}( & \mu v_{\ec{z_1}}. & \const{C}( v_{\ec{z_2}} ) ) & 
\Val \ec{x} & = & v_{\ec{x}} \\
\end{array}\]

To construct a completion of $\Val$, we must choose values for the free variables of $\Val$, $v_{\ec{x}}$ and $v_{\ec{w_3}}$.
The content of the set $\mathcal{T}^\M_{v_{\ec{x}}}( \Val )$ is $\alpha$-equivalent to:
\[%\begin{array}{c}
\{
\mu x. \const{A}( \mu y. \const{B}( x )),
\mu x. \const{A}( \mu y. \const{B}( y )),
\mu x. \const{B}( x ),
\mu x. \const{C}( \mu y. \const{A}( x )),
\mu x. \const{A}( \mu y. \const{C}( x )),
\mu x. \const{A}( x )
\}
%\end{array}
\]
Now, consider a model $\M$ that interprets variables in $\Ec$ based on the mapping $\Val$,
for instance, $x$ is interpretted as $\Val \ec{x}$, $y_1$ as $\Val \ec{y_1}$ and so on.
Assigning a value for $v_{\ec{x}}$ that is $\alpha$-equivalent to a value in $\mathcal{T}^\M_{v_{\ec{x}}}( \Val )$
may cause values in the range of $\Val$ to become $\alpha$-equivalent, 
which in turn may cause $\M$ to be inconsistent.
For instance, say we assign $\mu v_{\ec{x}}. \const{B}( v_{\ec{x}} )$ for $v_{\ec{x}}$.
After this substitution, $\Val \ec{y_2}$ is $\mu v_{\ec{y_2}}. \const{B}( \mu v_{\ec{x}}. \const{B}( v_{\ec{x}} ) )$, 
which has normal form $\mu v_{\ec{y_2}}. \const{B}( v_{\ec{y_2}} )$,
which is $\alpha$-equivalent to $\mu v_{\ec{x}}. \const{B}( v_{\ec{x}} )$.
However, this contradicts the disequality $x \not\teq y_2$ in $\Ec$.
Inversely, if the value assigned to $v_{\ec{x}}$ is not $\alpha$-equivalent to any term in $\mathcal{T}^\M_{v_{\ec{x}}}( \Val )$,
then the values in the range of $\Val$ remain $\alpha$-disequivalent.
For instance, we may assign values such as 
$\mu v_{\ec{x}}. \const{C}( v_{\ec{x}} )$,
$\mu v_{\ec{x}}. \const{C}( \mu x_1. \const{A}( x_1) )$, and
$\mu v_{\ec{x}}. \const{B}( \mu x_1. \const{A}( v_{\ec{x}} ) )$
to $v_{\ec{x}}$.
Notice that legal substitutions for $v_{\ec{x}}$ may cause the range of $\Val$ to contain non-normal terms,
as in after applying the third substitution mentioned above,
$\Val \ec{w_1}$ is $\mu v_{\ec{w_1}}. \const{A}( \mu v_{\ec{x}}. \const{B}( \mu x_1. \const{A}( v_{\ec{x}} ) ) )$.
$\Box$
\end{example}

@@

\begin{lemma} 
\label{lem:model-completion}
If $\Val$ is constructed for a saturated $\Ec$,
and $\Val^\ast$ is a completion of $\Val$ for normal model $\M$, then:
\begin{enumerate}
\item $\Val^\ast \ec{x}$ is bisimilar with a value in $\M( \tau )$, for all $\ec{x}$ of type $\tau$ of $\Ec$,
\item
If $\const{C}( t_1, \ldots, t_n ) \in \ec{x}$,
then $\Val^\ast \ec{x} = \mu v_{\ec{x}}. \const{C}( w_1, \ldots, w_n )$
and $\Val^\ast \ec{x} \interp{ w_i } = \Val^\ast \ec{t_i}$ for $i = 1, \ldots, n$.
%and $\muvar( w_i ) = v_{\ec{t_i}}$ for $i = 1, \ldots, n$,
%\item
%$t_1 \interp{ t_2 } = \Val^\ast \ec{x}$
%for all $t_1$ with subterm $t_2$ in the range of $\Val^\ast$ where $\muvar(t_2)=v_{\ec{x}}$,
\item 
$\Val^\ast \ec{x} \vsim \Val^\ast \ec{y}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}
\begin{proof}
To show 1, we first show that $\Val^\ast$ contains no free variables.
Assume by contradiction that $\Val^\ast$ contained a free variable $v_{\ec{y}}$ for some $\ec{y}$ of type $\tau$.
Then it must be the case that $\ec{y}$ does not contain a constructor term,
or else $v_{\ec{y}}$ would not occur as a free variable in $\Val$.
Consider the case when $\tau$ is finite.
By assumption, $\tau \not\in \Nondata$.
Since \rn{Split} does not apply to $\Ec$, we have $\tau \not\in \Data$.
If $\tau \in \Codata$, then $\tau$ is corecursive by assumption, and by Lemma~\ref{lem:corecursive-singletons},
the cardinality of $\tau$ must be one.
Since \rn{Singleton} does not apply to $\Ec$,
there is only one equivalence class of type $\tau$ in $\Ec$,
and thus there are no terms in $\mathcal{T}^\M_x(\Val^\ast)$ of type $\tau$.
This is a contradiction, since our model completion may assign the value in the domain of $\tau$ to $v_{\ec{x}}$.
Now, consider the case when $\tau$ is infinite.
This is also a contradiction, 
since there are only a finite number of closed terms in $\mathcal{T}^\M_x(\Val^\ast)$,
and thus our model completion can assign a value not occurring in $\mathcal{T}^\M_x(\Val^\ast)$ to $v_{\ec{y}}$.
By construction, $\Val^\ast \ec{x}$ is normal.
Since \rn{Acyclic} does not apply, $\Val \ec{x}$ is acyclic when $\tau \in \Data$.
Moreover, our construction of $\Val^\ast$ applies substitutions of the form 
$\{ y \mapsto t \}$, where $t$ is acyclic when the type of $y$ is not a codatatype.
Thus, $\Val^\ast \ec{x}$ is acyclic when $\tau \not\in \Codata$.
Therefore, by definition, $\Val^\ast \ec{x}$ is bisimilar with a value in $\M( \tau )$.

>>>>

%\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\vec}[1]{\bar #1}
\newcommand{\Ec}{\mathsf{E}}
\newcommand{\Fc}{\mathsf{F}}
\newcommand{\Gc}{\mathsf{G}}
\newcommand{\Ac}{\mathsf{A}}
\newcommand{\Dc}{\mathsf{D}}
\newcommand{\Rc}{\mathsf{R}}
\newcommand{\vrange}{\mathsf{range}}
\newcommand{\vdom}{\mathsf{dom}}
\newcommand{\tEc}{\mathcal{T}(\Ec)}
\newcommand{\tcEc}{\mathcal{T}^\ast(\Ec)}
\newcommand{\rn}[1]{\textsf{\small #1}}
\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\tneq}{\not\teq}
\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
%\newcommand{\is}[1]{is\text{-} #1}
\newcommand{\is}[1]{\const{is#1}}
\newcommand{\ror}{\quad \parallel \quad}
\newcommand{\tpath}[2]{\mathcal{P}_{ #2 \rightarrow \_ }( #1 )}
\newcommand{\ttpath}[3]{\mathcal{P}_{ #2 \rightarrow #3 }( #1 )}
\newcommand{\interp}[1]{[\![ #1 ]\!]}
\newcommand{\dpath}[3]{\delta^{#1}_{#2}( #3 )}
\newcommand{\ec}[1]{ [ #1 ] }
\newcommand{\M}{\mathcal{M}}
\newcommand{\Val}{\mathcal{A}}
\newcommand{\Var}{\mathcal{V}}
\newcommand{\nf}[1]{{#1}\!\!\downarrow}
\newcommand{\vsim}{\sim}
\newcommand{\aequiv}{\mathrel{=_\alpha}}
\newcommand{\vsimv}[1]{\vsim_{#1}}
\newcommand{\tpos}[2]{#1\!\mid_{#2}}
\newcommand{\muvar}{\mathcal{Var}}

@@

\paragraph{Phase 2: Applying Acyclicity and Bisimilarity.}
%For presentation of the rules in this phase, we rely on a representation of terms in the $\mu$-notation.
The rules of the second phase contain premises that involve a mapping $\Val$
that associates a $\mu$-term as representative with each equivalence class.
%Recall that $\mu$-bindings can be used for representing possibly cyclic terms and values.
%(Section~\ref{sec:introduction}).
% for example, $\mu x. \const{C}( \const{0}, x )$ represents the cyclic value $\const{C}( \const{0}, \const{C}( \const{0}, \ldots ))$.
A $\mu$-term $t$ of type $\tau$ describes a class of values of
$\tau$ that terms in that equivalence class can take in the models of $\Ec$;
%either in the case when $\tau$ is a codatatype or datatype type,
when $\tau \in \Data$, a cyclic $t$ describes an infeasible class of values.
%$\tau \not\in \Codata$.
Formally, $\mu$-terms are defined recursively as being either a variable $x$
or a node of the form
$\mu x.\; \const{C}( \vec t\, )$ for some constructor $\const{C} \in \Ctr$ and
$\mu$-terms $\vec t$ with the expected types.
The variable $x$ need not occur freely under the binder's body, in which case the
binder can be omitted.
In the \rn{Acyclic} rule,
$\FV( u )$ denotes the variables occurring freely in $u$.

@@

The mapping $\Val$ is computed as follows.
%
For each equivalence class $\ec{x}$, we associate a fresh variable $\Var{\ec{x}}$ of the same type as $x$
%not occurring in $\tEc$,
and let $\Val\ec{x} = \Var{\ec{x}}$.
Thus, there are initially no constraints on the values for any equivalence class $\ec{x}$.
The mapping $\Val$ is then modified by applying the following rule exhaustively:
\[
\hbox{\(
\inferrule{
  \Var{\ec{x}} \in \FV( \Val )
  \quad
  \const{C}( t_1, \ldots, t_n ) \in \ec{x}
  \quad
  \const{C} \in \Ctr
}{
  \Val := \Val \{ \Var{\ec{x}} \mapsto \mu \Var{\ec{x}}.\; \const{C}( \Var{\ec{t_1}}, \ldots, \Var{\ec{t_n}} ) \}
}
\)
\rn{Update}}
\]
In the rule, $\FV( \Val )$ denotes the union of the free variables occurring in $\Val$'s range.
It is straightforward to see that the height of terms produced as a result of this update
is bounded by the number of equivalence classes of $\Ec$,
and thus our construction of $\Val$ is terminating.

\begin{example}
Say that $\Ec$ contains (distinct) equivalence classes $\ec{x}$, $\ec{y}$, $\ec{z}$ and $\ec{w}$ 
where $\const{C}(z,x) \in \ec{y}$ and $\const{C}(w,y) \in \ec{x}$ for some $\const{C} \in \Ctr$.
A possible sequence of updates applied to $\Val$ is shown below.
In the first column, we show the equivalence class that was considered,
and in the second column, we show the content of $\Val$.
\[\begin{array}{|l|@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l|}
\hline
\text{Eqc} & \multicolumn{3}{c}{\Val} & \ 
\\
\hline
& \{ & \ec{x} \mapsto \Var{\ec{x}}, & \ec{z} \mapsto \Var{\ec{z}}, & \\
&    & \ec{y} \mapsto \Var{\ec{y}}, & \ec{w} \mapsto \Var{\ec{w}} & \} \\
\ec{x}
& \{ & \ec{x} \mapsto \mu \Var{\ec{x}}. \const{C}( \Var{\ec{w}}, \Var{\ec{y}} ), & \ec{z} \mapsto \Var{\ec{z}}, & \\
&    & \ec{y} \mapsto \Var{\ec{y}}, & \ec{w} \mapsto \Var{\ec{w}} & \} \\
\ec{y}
& \{ & \ec{x} \mapsto \mu \Var{\ec{x}}. \const{C}( \Var{\ec{w}}, \mu \Var{\ec{y}}. \const{C}( \Var{\ec{z}}, \Var{\ec{x}} ) ), & \ec{z} \mapsto \Var{\ec{z}}, & \\
&    & \ec{y} \mapsto \mu \Var{\ec{y}}. \const{C}( \Var{\ec{z}}, \Var{\ec{x}} ), & \ec{w} \mapsto \Var{\ec{w}} & \} \\
\ec{x}
& \{ & \ec{x} \mapsto \mu \Var{\ec{x}}. \const{C}( \Var{\ec{w}}, \mu \Var{\ec{y}}. \const{C}( \Var{\ec{z}}, \Var{\ec{x}} ) ), & \ec{z} \mapsto \Var{\ec{z}}, & \\
&    & \ec{y} \mapsto \mu \Var{\ec{y}}. \const{C}( \Var{\ec{z}}, \mu \Var{\ec{x}}. \const{C}( \Var{\ec{w}}, \Var{\ec{y}} ) ), & \ec{w} \mapsto \Var{\ec{w}} & \} \\
\hline
\end{array}\]
This indicates that the values for $x$ (resp. $y$) in models of $\Ec$
are of the form $\const{C}( \Var{\ec{w}}, \const{C}( \Var{\ec{z}}, \ldots ))$ (resp. $\const{C}( \Var{\ec{z}}, \const{C}( \Var{\ec{w}}, \ldots ))$) 
for some value of $\Var{\ec{w}}$ and $\Var{\ec{z}}$.
\qed
\end{example}

@@

\begin{example}
Let $\const{C}$ be a constructor of type $\tau_1 \times \tau_2 \rightarrow \tau_2$,
and let $\Ec$ be the set $\{ x \teq \const{C}(y,\const{C}(y,x)) \}$.
After applying the rules in Figure~\ref{fig:cc-rules} to completion, the equivalence classes of $\Ec$ are
$\{ x, \const{C}(y,\const{C}(y,x)) \}$, $\{ \const{C}(y,x) \}$, and $\{ y \}$.
We then construct $\Val$ using the aforementioned phases, after which we obtain:
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}r@{\hspace{.5em}}l@{\hspace{.5em}}r@{\hspace{.5em}}l}
\Val \ec{x} & = & 
\mu \Var{\ec{x}}. & \const{C}( \Var{\ec{y}}, & \mu \Var{\ec{\const{C}(y,x)}}. & \const{C}( \Var{\ec{y}}, \Var{\ec{x}} ) ) \\
\Val \ec{\const{C}(y,x)} & = & 
\mu \Var{\ec{\const{C}(y,x)}}. & \const{C}( \Var{\ec{y}}, & \mu \Var{\ec{x}}. & \const{C}( \Var{\ec{y}}, \Var{\ec{\const{C}(y,x)}} ) )
\end{array}\]
Notice that $\Val \ec{x}$ and $\Val \ec{\const{C}(y,x)}$ are syntactically equivalent for the renaming
$\{ \Var{\ec{x}} \mapsto \Var{\ec{\const{C}(y,x)}}, \Var{\ec{\const{C}(y,x)}} \mapsto \Var{\ec{x}} \}$,
and thus our calculus may infer $x \teq \const{C}(y,x)$.
Intuitively, this equality holds, since the values of $x$ and $\const{C}( y, x )$ 
are equivalent for any value of $\Var{\ec{y}}$.
\qed
\end{example}

@@

\begin{example}
Let $\tau$ be a type in $\Codata$ with three constructors $\const{A}$, $\const{B}$ and $\const{C}$ each having type $\tau \rightarrow \tau$.
Let $\Ec$ be the set consisting of equalities
$\{$
$y_1 \teq \const{A}( y_2 )$,
$y_2 \teq \const{B}( x )$,
$z_1 \teq \const{C}( z_2 )$,
$z_2 \teq \const{A}( z_1 )$,
$w_1 \teq \const{A}( x )$,
$w_2 \teq \const{C}( w_3 )$,
$x \not\teq y_2$,
$\}$.
After running the calculus to saturation on $\Ec$, the mapping $\Val$ contains:
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{2.0em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
\Val \ec{y_1} & = & \mu \Var{\ec{y_1}}. & \const{A}( & \mu \Var{\ec{y_2}}. & \const{B}( \Var{\ec{x}} ) ) &
\Val \ec{w_1} & = & \mu \Var{\ec{w_1}}. & \const{A}( \Var{\ec{x}} )      & \\ 
\Val \ec{y_2} & = & \mu \Var{\ec{y_2}}. & \multicolumn{2}{l}{\const{B}( \Var{\ec{x}} )}      & &
\Val \ec{w_2} & = & \mu \Var{\ec{w_2}}. & \const{C}( \Var{\ec{w_3}} ) \\
\Val \ec{z_1} & = & \mu \Var{\ec{z_1}}. & \const{C}( & \mu \Var{\ec{z_2}}. & \const{A}( \Var{\ec{z_1}} ) ) &
\Val \ec{w_3} & = & \Var{\ec{w_3}} \\
\Val \ec{z_2} & = & \mu \Var{\ec{z_2}}. & \const{A}( & \mu \Var{\ec{z_1}}. & \const{C}( \Var{\ec{z_2}} ) ) & 
\Val \ec{x} & = & \Var{\ec{x}} \\
\end{array}\]

To construct a completion of $\Val$, we must choose values for the free variables of $\Val$, $\Var{\ec{x}}$ and $\Var{\ec{w_3}}$.
The content of the set $\mathcal{T}^\M_{\Var{\ec{x}}}( \Val )$ is $\alpha$-equivalent to:
\[%\begin{array}{c}
\{
\mu x. \const{A}( \mu y. \const{B}( x )),
\mu x. \const{A}( \mu y. \const{B}( y )),
\mu x. \const{B}( x ),
\mu x. \const{C}( \mu y. \const{A}( x )),
\mu x. \const{A}( \mu y. \const{C}( x )),
\mu x. \const{A}( x )
\}
%\end{array}
\]
Now, consider a model $\M$ that interprets variables in $\Ec$ based on the mapping $\Val$,
for instance, $x$ is interpretted as $\Val \ec{x}$, $y_1$ as $\Val \ec{y_1}$ and so on.
Assigning a value for $\Var{\ec{x}}$ that is $\alpha$-equivalent to a value in $\mathcal{T}^\M_{\Var{\ec{x}}}( \Val )$
may cause values in the range of $\Val$ to become $\alpha$-equivalent, 
which in turn may cause $\M$ to be inconsistent.
For instance, say we assign $\mu \Var{\ec{x}}. \const{B}( \Var{\ec{x}} )$ for $\Var{\ec{x}}$.
After this substitution, $\Val \ec{y_2}$ is $\mu \Var{\ec{y_2}}. \const{B}( \mu \Var{\ec{x}}. \const{B}( \Var{\ec{x}} ) )$, 
which has normal form $\mu \Var{\ec{y_2}}. \const{B}( \Var{\ec{y_2}} )$,
which is $\alpha$-equivalent to $\mu \Var{\ec{x}}. \const{B}( \Var{\ec{x}} )$.
However, this contradicts the disequality $x \not\teq y_2$ in $\Ec$.
Inversely, if the value assigned to $\Var{\ec{x}}$ is not $\alpha$-equivalent to any term in $\mathcal{T}^\M_{\Var{\ec{x}}}( \Val )$,
then the values in the range of $\Val$ remain $\alpha$-disequivalent.
For instance, we may assign values such as 
$\mu \Var{\ec{x}}. \const{C}( \Var{\ec{x}} )$,
$\mu \Var{\ec{x}}. \const{C}( \mu x_1. \const{A}( x_1) )$, and
$\mu \Var{\ec{x}}. \const{B}( \mu x_1. \const{A}( \Var{\ec{x}} ) )$
to $\Var{\ec{x}}$.
Notice that legal substitutions for $\Var{\ec{x}}$ may cause the range of $\Val$ to contain non-normal terms,
as in after applying the third substitution mentioned above,
$\Val \ec{w_1}$ is $\mu \Var{\ec{w_1}}. \const{A}( \mu \Var{\ec{x}}. \const{B}( \mu x_1. \const{A}( \Var{\ec{x}} ) ) )$.
$\Box$
\end{example}

@@

\begin{lemma} 
\label{lem:model-completion}
If $\Val$ is constructed for a saturated $\Ec$,
and $\Val^\ast$ is a completion of $\Val$ for normal model $\M$, then:
\begin{enumerate}
\item $\Val^\ast \ec{x}$ is bisimilar with a value in $\M( \tau )$, for all $\ec{x}$ of type $\tau$ of $\Ec$,
\item
If $\const{C}( t_1, \ldots, t_n ) \in \ec{x}$,
then $\Val^\ast \ec{x} = \mu \Var{\ec{x}}. \const{C}( w_1, \ldots, w_n )$
and $\Val^\ast \ec{x} \interp{ w_i } = \Val^\ast \ec{t_i}$ for $i = 1, \ldots, n$.
%and $\muvar( w_i ) = \Var{\ec{t_i}}$ for $i = 1, \ldots, n$,
%\item
%$t_1 \interp{ t_2 } = \Val^\ast \ec{x}$
%for all $t_1$ with subterm $t_2$ in the range of $\Val^\ast$ where $\muvar(t_2)=\Var{\ec{x}}$,
\item 
$\Val^\ast \ec{x} \vsim \Val^\ast \ec{y}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}
\begin{proof}
To show 1, we first show that $\Val^\ast$ contains no free variables.
Assume by contradiction that $\Val^\ast$ contained a free variable $\Var{\ec{y}}$ for some $\ec{y}$ of type $\tau$.
Then it must be the case that $\ec{y}$ does not contain a constructor term,
or else $\Var{\ec{y}}$ would not occur as a free variable in $\Val$.
Consider the case when $\tau$ is finite.
By assumption, $\tau \not\in \Nondata$.
Since \rn{Split} does not apply to $\Ec$, we have $\tau \not\in \Data$.
If $\tau \in \Codata$, then $\tau$ is corecursive by assumption, and by Lemma~\ref{lem:corecursive-singletons},
the cardinality of $\tau$ must be one.
Since \rn{Singleton} does not apply to $\Ec$,
there is only one equivalence class of type $\tau$ in $\Ec$,
and thus there are no terms in $\mathcal{T}^\M_x(\Val^\ast)$ of type $\tau$.
This is a contradiction, since our model completion may assign the value in the domain of $\tau$ to $\Var{\ec{x}}$.
Now, consider the case when $\tau$ is infinite.
This is also a contradiction, 
since there are only a finite number of closed terms in $\mathcal{T}^\M_x(\Val^\ast)$,
and thus our model completion can assign a value not occurring in $\mathcal{T}^\M_x(\Val^\ast)$ to $\Var{\ec{y}}$.
By construction, $\Val^\ast \ec{x}$ is normal.
Since \rn{Acyclic} does not apply, $\Val \ec{x}$ is acyclic when $\tau \in \Data$.
Moreover, our construction of $\Val^\ast$ applies substitutions of the form 
$\{ y \mapsto t \}$, where $t$ is acyclic when the type of $y$ is not a codatatype.
Thus, $\Val^\ast \ec{x}$ is acyclic when $\tau \not\in \Codata$.
Therefore, by definition, $\Val^\ast \ec{x}$ is bisimilar with a value in $\M( \tau )$.
